\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{indentfirst} % recua também o 1º parágrafo de cada seção (uniformidade)
\usepackage{hyperref}

\title{When the Baseline Also Abstains:\\
Pitfalls in Evaluating Abstention for Document Audit}
\author{Marcos de Paula \\ FLARE26 Project \\
\small ORCID: \href{https://orcid.org/0009-0001-9105-6100}{0009-0001-9105-6100}}
\date{Preprint --- June 2026}

\begin{document}
\maketitle

\begin{abstract}
Retrieval-augmented (RAG) auditors invent disagreements: asked for a value the
document never gives, they answer anyway, producing \textbf{false-positive
divergences}. We build an \textbf{ontology-gated} auditor that abstains on
type/scope mismatch---and, more importantly, we find that \emph{measuring}
whether such a system abstains well is itself error-prone. Using a benchmark of
Brazilian contracts and government tender notices, we surface \textbf{three
pitfalls}, each illustrated by a concrete mistake we made and corrected: (i) a
\textbf{single-sample illusion}, where a rare false-positive rate reads as
``0\%'' until bootstrapped, and a self-consistency ``free lunch'' turns out to be
sampling noise; (ii) that \textbf{``hard'' is defined by the model, not the
taxonomy}---a naive baseline ties an abstaining system not only on wholly-missing
concepts but even on taxonomically-adjacent ones (an engineering benchmark we
built to discriminate did not), so a gate helps only where the model actually
conflates a concept pair; and (iii) a \textbf{keyword-absence annotation trap},
where labeling \texttt{ABSTAIN} from a missing keyword is invalid on real legal
text that paraphrases concepts. We distill the lessons into a checklist for
evaluating abstention in document audit, and release the system, benchmark, and
harnesses.
\end{abstract}

\section{Introduction}
Organizations compare many structured-but-unstructured documents---contracts,
tender notices (\emph{editais}), financial statements, engineering specification
sheets---to find discrepancies. A
natural approach is RAG: retrieve passages and ask an LLM to extract the answer
per document, then compare. This fails in a specific, costly way:
\textbf{standard RAG prefers answering to abstaining.} Asked ``what is the
penalty for \emph{total non-performance}?'' of a contract that only sets a
\emph{late-payment} penalty, a naive extractor returns the late-payment
value---fabricating an answer to a question the document never addresses. Across
documents these become \textbf{false-positive divergences}: flagged
disagreements that are artifacts of hallucination, and in auditing the costliest
error.

The obvious fix is to make the system abstain. We do (Section~\ref{sec:system}),
and evaluate it across \textbf{two domains}---Brazilian legal contracts and tender
notices (Portuguese), and English reliability and mechanical engineering
specification sheets---so the lessons do not hinge on one domain.
But our main finding is methodological: \textbf{whether an abstaining system is
actually better is surprisingly hard to measure}, and the natural ways to measure
it quietly mislead. We report three pitfalls, each a mistake we made first and
caught later:
\begin{itemize}
  \item \textbf{Pitfall A --- single-sample illusion} (\S\ref{sec:pa}). Our first
  runs showed ``$0\% \pm 0\%$'' false positives; bootstrapping revealed
  $\sim$1--2\%. The zero was sampling noise, and it made a permissive
  self-consistency vote look like free recall.
  \item \textbf{Pitfall B --- ``hard'' is the model's, not the taxonomy's}
  (\S\ref{sec:pb}). A baseline ties our gate not only where a concept is wholly
  absent but even on taxonomically-adjacent pairs (a 21-case engineering
  benchmark we built to discriminate did not). A gate helps only where the
  \emph{model} conflates a pair.
  \item \textbf{Pitfall C --- keyword-absence annotation trap}
  (\S\ref{sec:pc}). We labeled \texttt{ABSTAIN} from a missing keyword; three
  labels were wrong because the concept was present under another name. The gate
  had answered correctly; our benchmark was lying.
\end{itemize}

\paragraph{Contributions.} (1) An ontology-gated auditor with explicit abstention
and a deterministic comparison judge (case study, Section~\ref{sec:system}). (2)
Three evaluation pitfalls for abstention in document audit, each demonstrated on
a real failure (Section~\ref{sec:eval}). (3) A checklist for evaluating
abstention, plus a released benchmark and harnesses.\footnote{Code, benchmark,
and evaluation harnesses: \url{https://github.com/marcospaula/FLARE26-P1}}

\section{Related Work}
\textbf{RAG and its hallucinations} \cite{lewis2020rag,gao2023ragsurvey}; our
concern is answering despite absent in-scope evidence. \textbf{Structured
extraction} (function-calling, Instructor \cite{instructor}, Guardrails
\cite{guardrails}, constrained decoding \cite{willard2023guided}) coerces typed
outputs; we make the schema's primary job an abstention pre-condition.
\textbf{Selective prediction / abstention}
\cite{geifman2017selective,kadavath2022know} and answerability
\cite{rajpurkar2018squad2} are the closest neighbors; we study how to
\emph{evaluate} abstention in multi-document audit. \textbf{Faithfulness and RAG
evaluation} \cite{ji2023survey,maynez2020faithfulness,es2024ragas} ask whether an
answer is supported; we target the dual question---whether a system correctly
declines---and how its benchmarks mislead. \textbf{Legal NLP}
\cite{hendrycks2021cuad,chalkidis2020legalbert} supplies the domain.

\section{The Case-Study System}
\label{sec:system}
FLARE26 is a RAG pipeline used as a vehicle for the evaluation study: hybrid
retrieval (M1.5) $\to$ ontology-gated extraction (M2) $\to$ deterministic
$N$-way judge (M4) $\to$ summary (M5), with provenance to the source span.

\textbf{Ontology-gated extraction (M2).} A typed record with two independent
checks: \emph{type/instrument} (the named instrument---e.g.\ a \emph{penalty}
vs.\ \emph{interest}, distinct even when both are percentages; must match) and
\emph{scope/condition} (strict equality). If either fails, the record collapses
to an explicit \emph{evidence gap}. The rule is \textbf{domain-agnostic} (stated
over abstract instruments $P/Q$, conditions $A/B$; no hard-coded legal terms).

\textbf{Deterministic $N$-way judge (M4).} Groups documents by equivalent answer
(BR number normalization: \texttt{R\$ 1.200.000,50} $\to$ \texttt{1200000.5};
\texttt{12,5\%} $\equiv$ \texttt{12.5}); verdict \emph{consensus} /
\emph{divergence} / \emph{evidence gap} in $O(N)$, no LLM call. An upstream
hallucination is the only way to fabricate a divergence---what the gate
suppresses. \textbf{Optional self-consistency:} run $k$ times and vote; the
threshold $t$ is a precision/recall knob (\S\ref{sec:pa}).

\section{Evaluation and Three Pitfalls}
\label{sec:eval}
\textbf{Benchmark.} 61 verified (question, document) pairs across two domains and
two languages: a \textbf{legal} split---30 synthetic Brazilian contracts (13
\texttt{ABSTAIN}, 17 answerable) and 10 real tender notices (4 \texttt{ABSTAIN},
6 answerable), in Portuguese---and an \textbf{engineering} split---21 English
reliability/mechanical spec sheets (12 \texttt{ABSTAIN}, 9 answerable), with both
\emph{type} near-neighbors (MTBF vs.\ MTTR, tensile vs.\ yield) and \emph{scope}
near-neighbors (same metric, different condition: tensile at 20\,$^\circ$C vs.\
200\,$^\circ$C; torque for bolt class 8.8 vs.\ 10.9). Extractor: gpt-4o-mini,
temperature $0$, fixed seed. Metrics: false-positive rate (fraction of
\texttt{ABSTAIN} answered), abstention recall, answer recall.

\subsection{The synthetic result (what we wanted to report)}
\begin{table}[h]\centering
\begin{tabular}{lcc}
\toprule
System & False-positive $\downarrow$ & Answer recall $\uparrow$ \\
\midrule
Baseline (free extraction) & $38\% \pm 0\%$ & $100\% \pm 0\%$ \\
FLARE26 (gated)            & $\sim$1--2\%   & $81\% \pm 6\%$ \\
\bottomrule
\end{tabular}
\end{table}
Ontological gating cuts false-positive divergences $\sim$20--30$\times$. Ablation:
relaxing the strict scope rule to accept sub/superset conditions recovers recall
but \textbf{reopens} false positives ($0\% \to 38\%$), so the strict rule is the
operating point.

\subsection{Pitfall A --- the single-sample illusion}
\label{sec:pa}
Our first measurement reported \textbf{$0\% \pm 0\%$} false positives, and a vote
``answer if $\geq 1$ of $k$'' appeared to recover recall \emph{for free}. Both
were artifacts of measuring a rare event once. Bootstrapping the per-item sample
pool:
\begin{table}[h]\centering
\begin{tabular}{lcc}
\toprule
$k$ (LLM calls/doc) & False-positive & Answer recall \\
\midrule
1  & $1\% \pm 2\%$ & $81\% \pm 6\%$ \\
5  & $3\% \pm 4\%$ & $91\% \pm 3\%$ \\
10 & $5\% \pm 4\%$ & $93\% \pm 2\%$ \\
\bottomrule
\end{tabular}
\end{table}
The gate leaks rarely ($\sim$1--2\% per sample), and ``$\geq 1$'' \emph{amplifies}
it with $k$. Sweeping the vote threshold $t$ shows ``$\geq 1$'' was the wrong
knob \cite{wang2023selfconsistency}; a majority-style $t$ filters the leaks:
\begin{table}[h]\centering
\begin{tabular}{lcc}
\toprule
Policy & False-positive & Answer recall \\
\midrule
single call ($k{=}1$)      & $2\%$  & $77\%$ \\
$k{=}10, t{=}1$ (``$\geq 1$'') & $10\%$ & $86\%$ \\
$k{=}10, t{=}3$            & $1\%$  & $82\%$ \\
$k{=}10, t{=}4$           & $\mathbf{0\%}$ & $\mathbf{80\%}$ \\
\bottomrule
\end{tabular}
\end{table}
(The two tables are independent bootstrap runs; the small disagreement at
$k{=}1$---$1\%$ vs.\ $2\%$ FP, $81\%$ vs.\ $77\%$ recall---is itself the sampling
noise this section is about.) The point $k{=}10, t{=}4$ dominates the single call
on both axes. \textbf{Lesson:} report rare-event metrics as mean $\pm$ std over
repeated runs; sweep the vote threshold; never trust a single ``$0\%$''.

\subsection{Pitfall B --- ``hard'' is defined by the model, not the taxonomy}
\label{sec:pb}
The legal $38\%$ baseline false-positive rate comes from absences where a
\emph{similar-but-wrong concept is present}---asked for a \emph{penalty}, the
document has \emph{interest}. It is tempting to call these ``hard'' because the
concepts are taxonomically adjacent; they are hard rather because \textbf{the
model conflates the pair}. Two controls make this precise.

\textbf{Wholly-absent concepts are easy---even for the baseline.} Our first real
\texttt{ABSTAIN} cases had the concept wholly missing; retrieval returns nothing
and \textbf{even the baseline abstains}, tying the gate at $0\%$. A 10-pair real
pilot (4 such absences, 6 answerable) gives FLARE $0\%$ / $83\%$ recall vs.\
baseline $0\%$ / $50\%$: the gate wins on recall, the false-positive comparison is
uninformative.

\textbf{Taxonomically-adjacent $\neq$ hard.} We built 21 engineering cases with
\emph{type} near-neighbors (MTBF vs.\ MTTR, tensile vs.\ yield) and \emph{scope}
near-neighbors (20\,$^\circ$C vs.\ 200\,$^\circ$C; class 8.8 vs.\ 10.9). We
expected the baseline to leak; \textbf{it did not}---on all 12 engineering
\texttt{ABSTAIN} cases it abstained correctly ($0\%$ false positives), matching
the gate. The model does not confuse these crisp concepts. A control isolates the
cause: the legal \emph{interest-for-penalty} leak persists when the case is
translated to English, so it is the \textbf{concept pair}, not the language or
domain, that the model treats as interchangeable.

\textbf{Consequences.} An abstention gate helps \emph{only where the baseline
leaks}, which depends on the model's concept-confusability---not the annotator's
taxonomy, the domain, or the language. In crisp domains the baseline is already
careful, so the gate does no harm (the engineering split: $0\%$ false positives,
$100\%$ recall) but adds nothing. \textbf{Lesson:} establish concept-difficulty
empirically (does the baseline leak?), not by assuming adjacency implies it.

\subsection{Pitfall C --- the keyword-absence annotation trap}
\label{sec:pc}
We labeled real-document \texttt{ABSTAIN} from a missing \emph{keyword}. Three
labels were wrong, because legal text paraphrases concepts: \emph{juros de mora}
appeared as \emph{compensação financeira}; ``payment term'' as ``payment within
30 days''. In each case the gate had \textbf{answered correctly} and our label
called it a false positive. Re-annotating absence by \textbf{concept} (exhausting
synonyms before declaring \texttt{ABSTAIN}) corrected them; one ambiguous case is
held out pending domain adjudication. \textbf{Lesson:} never infer absence from a
missing keyword on real text; annotate by concept, with a second annotator.

\subsection{Separating retrieval from the gate}
\label{sec:pe}
Diagnosing a real-document recall failure revealed a \textbf{retrieval} bug, not a
gate failure: the context assembler truncated an unordered set, occasionally
dropping the highest-ranked lexical block (the exact penalty clause). A
relevance-ordered merge raised real-document recall from $67\%$ to $83\%$ at
unchanged false positives. \textbf{Lesson:} attribute failures to retrieval vs.\
extraction before blaming the gate.

\section{Recommendations: a checklist for evaluating abstention}
\begin{enumerate}
  \item \textbf{Repeat and bootstrap.} Report false positives as mean $\pm$ std
  over $k$ runs; a single ``$0\%$'' on a rare event is an illusion.
  \item \textbf{Establish difficulty empirically.} Cover cases where a
  similar-but-wrong concept is present, but treat an absence as ``hard'' only
  where the baseline actually leaks---adjacency does not imply difficulty; report
  easy vs.\ hard.
  \item \textbf{Annotate by concept, not keyword.} Confirm absence by exhausting
  paraphrases; prefer a second annotator on real text.
  \item \textbf{Sweep the vote threshold.} ``Answer if any'' inflates false
  positives; report the $(k,t)$ operating curve.
  \item \textbf{Attribute failures.} Separate retrieval misses from gate
  decisions before drawing conclusions.
\end{enumerate}

\section{Discussion and Limitations}
\textbf{Small, partly synthetic benchmark:} a $\sim$1--2\% false-positive rate
cannot be estimated tightly from 13--17 \texttt{ABSTAIN} items. \textbf{Real-doc
external validity is open:} the real \texttt{ABSTAIN} cases are easy; hard real
cases coincide with the ambiguous ones awaiting adjudication---the priority
extension. \textbf{Single annotator; proprietary LLM:} no inter-annotator
agreement yet; open-model replication is future work. The system is an
\textbf{architectural pattern}, not a new ML method; the weight is on the
evaluation lessons and released artifacts.

\section{Conclusion}
An ontology-gated auditor does cut false-positive divergences $\sim$20--30$\times$
on a synthetic benchmark. But the more durable contribution is a caution:
evaluating abstention in document audit is easy to get wrong. A rare
false-positive rate looks like zero until you bootstrap; a naive baseline ties an
abstaining system not only on wholly-missing concepts but on taxonomically-adjacent
ones too---so ``hard'' is the model's to define, not the taxonomy's, and a gate
helps only where the model actually conflates a pair; and labeling absence by
keyword silently corrupts a benchmark on real legal text. We hope the checklist, benchmark, and
harnesses help others measure abstention honestly. Future work: hard
real-document absences, a second annotator, and open-model replication.

% NOTE: bibliographic details verified against official sources (ACL Anthology,
% NeurIPS Proceedings, ACM DL, arXiv) on 2026-06-24. BibTeX in references.bib.
\begin{thebibliography}{99}
\bibitem{chalkidis2020legalbert} Chalkidis, I., Fergadiotis, M., Malakasiotis, P., Aletras, N., \& Androutsopoulos, I. (2020). LEGAL-BERT: The Muppets straight out of Law School. \emph{Findings of EMNLP}.
\bibitem{es2024ragas} Es, S., James, J., Espinosa-Anke, L., \& Schockaert, S. (2024). RAGAS: Automated Evaluation of Retrieval Augmented Generation. \emph{EACL (System Demonstrations)}.
\bibitem{gao2023ragsurvey} Gao, Y., Xiong, Y., Gao, X., et al. (2023). Retrieval-Augmented Generation for Large Language Models: A Survey. \emph{arXiv:2312.10997}.
\bibitem{geifman2017selective} Geifman, Y., \& El-Yaniv, R. (2017). Selective Classification for Deep Neural Networks. \emph{NeurIPS}.
\bibitem{guardrails} Guardrails AI (2023). \emph{Guardrails}. Software. \url{https://github.com/guardrails-ai/guardrails}
\bibitem{hendrycks2021cuad} Hendrycks, D., Burns, C., Chen, A., \& Ball, S. (2021). CUAD: An Expert-Annotated NLP Dataset for Legal Contract Review. \emph{NeurIPS Datasets and Benchmarks}.
\bibitem{instructor} Liu, J. (2023). \emph{Instructor: Structured Outputs for LLMs}. Software. \url{https://github.com/567-labs/instructor}
\bibitem{ji2023survey} Ji, Z., Lee, N., Frieske, R., et al. (2023). Survey of Hallucination in Natural Language Generation. \emph{ACM Computing Surveys}.
\bibitem{kadavath2022know} Kadavath, S., Conerly, T., Askell, A., et al. (2022). Language Models (Mostly) Know What They Know. \emph{arXiv:2207.05221}.
\bibitem{lewis2020rag} Lewis, P., Perez, E., Piktus, A., et al. (2020). Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks. \emph{NeurIPS}.
\bibitem{maynez2020faithfulness} Maynez, J., Narayan, S., Bohnet, B., \& McDonald, R. (2020). On Faithfulness and Factuality in Abstractive Summarization. \emph{ACL}.
\bibitem{rajpurkar2018squad2} Rajpurkar, P., Jia, R., \& Liang, P. (2018). Know What You Don't Know: Unanswerable Questions for SQuAD. \emph{ACL}.
\bibitem{wang2023selfconsistency} Wang, X., Wei, J., Schuurmans, D., et al. (2023). Self-Consistency Improves Chain of Thought Reasoning in Language Models. \emph{ICLR}.
\bibitem{willard2023guided} Willard, B. T., \& Louf, R. (2023). Efficient Guided Generation for Large Language Models. \emph{arXiv:2307.09702}.
\end{thebibliography}

\end{document}
